HW2
========================================================
First we get the data and clean the fields. This code chunk is based on an email to the course list by Raphael.

```{r}
library(GEOmetadb)
gds = getGEO("GSE40812",destdir=getwd())
gds = gds[[1]]
pdata = pData(gds)
monocyte.pdata = pdata[pdata$source_name_ch1=="Monocyte-derived Macrophage",]
monocyte.eset<-gds[,rownames(monocyte.pdata)]
#characteristic_ch1 indicates VL++ or VL--
#characteristic_ch1.2 indicates Mock or Poly IC
monocyte.pdata$HCV = gsub(".*: ", "", monocyte.pdata$characteristics_ch1)
monocyte.pdata$HCV = ifelse(monocyte.pdata$HCV=="Neg", "neg", "pos")
monocyte.pdata$treatment = tolower(gsub(".*: ", "", monocyte.pdata$characteristics_ch1.2))
monocyte.pdata$treat = ifelse(monocyte.pdata$treatment=="mock","mock","poly")
subjst = regexpr("_",monocyte.pdata$title)
monocyte.pdata$subject = substr(monocyte.pdata$title,subjst+1,subjst+4)
```

Now that the data are cleaned up a bit we perform the first analysis to determine which probes are differentially expressed according to treatment type.
```{r}
source("http://bioconductor.org/biocLite.R")
biocLite("limma")
library(limma)
library(Biobase)
library(data.table)

mm1 = model.matrix(~treat+HCV, monocyte.pdata)
fit1 = lmFit(monocyte.eset, mm1)
ebay1 = eBayes(fit1)

tests1 = decideTests(ebay1,method="separate",adjust.method="BH",p.value=0.05,lfc=log2(1.5))
diffexp = which(tests1[,"treatpoly"] != 0)
length(tests1[diffexp,1]) ## 1146
```

Next we subset the expression set to include only the probes that were differentially expressed in the first limma analysis. Then we run another model in limma that fits treatment and HCV status as 4 factor variables: Poly Ic/VL-, Poly Ic/VL+, Mock/VL-, Mock/VL+ and use contrasts between VL+ and VL- to find which probes are differentially expressed within the treatment categories.

```{r}
## which probes are differentially expressed for treatment
polyic = rownames(tests1[diffexp,])
exprs(monocyte.eset) = exprs(monocyte.eset)[polyic,]

dim(exprs(monocyte.eset)) # 1146 40

treatHCV = factor(paste(monocyte.pdata$treat,monocyte.pdata$HCV,sep=""))
mm = model.matrix(~0+treatHCV, monocyte.pdata)
fit_mm = lmFit(monocyte.eset,mm)

cm = makeContrasts(treatHCVmockneg-treatHCVmockpos,treatHCVpolyneg-treatHCVpolypos,levels=mm)

fit_cm = contrasts.fit(fit_mm,cm)
ebay_cm = eBayes(fit_cm)

top_cm = topTable(ebay_cm,number=Inf,sort.by="none")
keepers = which(top_cm$P.Value<.1) # 52 of them
```

We end up with 52 probes instead of 43. I'm not sure why this is. For now we will construct the heatmap for these 52 probes. The first step is to reorganize the subjects according to VL status and treatment type.

```{r}
library(pheatmap)
#Subset the probes we found in the second analysis
kept = exprs(monocyte.eset)[keepers,]

#Get the phenodata we want to use in the heatmap
sm_pdata = monocyte.pdata[,c("subject","treat","HCV")]
#Sort by poly/mock and then by VL-/VL+
sm_pdata = sm_pdata[order(sm_pdata[,"treat"],sm_pdata[,"HCV"]),]
kept=kept[,rownames(sm_pdata)] #sorted by treatment type
colnames(kept)=sm_pdata[,"subject"] #Assign subject ids
```

Now that the subjects are properly ordered, we can do the heatmap:

```{r fig.width=7, fig.height=6}
pheatmap(mat=kept,cluster_rows=FALSE,cluster_cols=FALSE)
```


